Retrofitting Effect Handlers onto OCaml

### 2020-11-03 retro-concurrency.tex
Updates

\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

@@

@@

Effect Handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful
\emph{operations} from their interpretations, which are defined by
\emph{handlers} of the effects. For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an effect |Read|, which is parameterised with an input channel of type
|in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming. A computation may perform many
different effects, which are interpreted by one or more enclosing handlers.

@@

Common Lisp Conditions~\cite{Conditions} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision. If
the error is non-fatal, Conditions provide a restart mechanism to
resume execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We can implement
this mechanism using effect handlers as shown below:

@@

On line 4, we declare an effect |DivBy0| which is parameterised by an integer,
which when performed returns a value. While interpreting the division
expression, if the denominator is 0, then we perform |DivBy0| effect with the
numerator value (line 13) so that the context may decide whether the error is
fatal or not, and resume with an appropriate value if the error is not fatal.
We can get the |iOS| semantics by wrapping the evaluator computation with the
following handler:

\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}

If the evaluator returns a value, we simply return that. In case the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
other hand, we can get |gsearch| semantics by enclosing the evaluator in the
following handler:

\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
  if v = 0 then continue k Error
  else if v > 0 then continue k (Inf P)
  else continue k (Inf N)
\end{lstlisting}

The computation |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

We can extend the calculator modularly with the ability to read user input
using the |Read| effect defined earlier as follows:

\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}

The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:

\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
	continue (get_ui_k ()) ()
\end{lstlisting}

\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

@@

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
is meaningless and hence, we would lose compatibility with tools that inspect
the program stack. Hence, we choose not to CPS translation. The alternative is
to use the call stack for implementing continuation. Farvardin et
al~\cite{Folklore} show that there is a spectrum of implementation strategies
for continuation with different trade-offs.

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
exception |Unhandled| at the point of |perform|. We leave the question of
effect safety for future work.

@@

We motivate our implementation strategy based on the following ideal
requirements:

\begin{enumerate}[label=R\arabic*]
  \item \textbf{Backwards compatibility} A well behaved OCaml program does not
    break under OCaml extended with effect handlers. OCaml code that does not
    use effect handlers will pay negligible performance cost.
  \item \textbf{Tool compatibility} OCaml with effect handlers produce
    well-formed backtraces, and remains compatible with program analysis tools
    such as debuggers and profilers that inspect the stack using DWARF unwind
    tables.
  \item \textbf{Effect handler efficiency} In order to support
    highly-concurrent applications, the program must accommodate millions of
    continuations in the heap. Installing effect handlers, capturing and
    resuming continuations must be fast.
  \item \textbf{Forwards compatibility} While this work does not present an
    effect system, the design must be compatible with a future effect system
    for OCaml. As a cornerstone of modularity, we also want blocking I/O code
    to \emph{transparently} be made asynchronous with the help of effect
    handlers.
\end{enumerate}

The need to host millions of active continuations in the heap rules out the use
of large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate stack. Additionally, we must
generate DWARF stack unwind tables for stacks that may be non-contiguous and
contain frames from different languages. In this work, we develop the compiler
and runtime support for implementing efficient effect handlers for OCaml.

Our work is also timely. WebAssembly~\ref{} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{} includes direct-style asynchronous
programming and structured concurrency. We believe that our design choices will
inform similar choices to be made in the other languages.

@@

\begin{itemize}
  \item the design and implementation of affine effect handlers for OCaml. Our
    design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{}.
  \item a formal operation semantics for the effect handler implementation in
    OCaml.
  \item extensive evaluation which shows that our implementation has negligible
    impact on code that does not use effect handlers, and serves as an
    efficient foundation for scalable concurrent and parallel programming.
\end{itemize}

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation ``Multicore OCaml'' to
distinguish it from stock OCaml. Multicore OCaml uses effect handlers as the
mechanism to express user-level concurrency.

The rest of the paper is organized as follows. We describe the details of the
OCaml program stack in the next section, and highlight the features that the
effect handler implementation must retain in order to provide backwards
compatibility. In \S\ref{sec:refine}, we refine the design further
focussing on the issues that come up when integrating effect handlers into a
mainstream language. \S\ref{sec:semantics} presents the formal operational
semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
compiler and the runtime system support for implementing effect handlers.
\S\ref{sec:eval} presents extensive performance evaluation of effect handlers
against the goals set in \S\ref{sec:req}. \S\ref{sec:related} and
\S\ref{sec:conc} present the related work and the conclusions.

@@

\begin{figure*}
\begin{minipage}{0.59\linewidth}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[numbers=left]
external ocaml_to_c : unit -> int = "ocaml_to_c"
exception E1
exception E2

let c_to_ocaml () = raise E1
let _ = Callback.register "c_to_ocaml" c_to_ocaml

let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c () with E2 -> 0
  with E1 -> 42;;

let _ = assert (omain () = 42)
    \end{lstlisting}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
  \end{minipage}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[language=c,numbers=left]
#include <caml/mlvalues.h>
#include <caml/callback.h>

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
  \begin{minipage}{0.39\linewidth}
    \centering
    \includegraphics[scale=0.46]{figures/stock_stack}
    \subcaption{Stack layout before raising exception \texttt{E1}.}
    \label{fig:stock_stack}
  \end{minipage}
  \caption{Program stack on stock OCaml.}
\end{figure*}

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack. To this end, in this section, we provide an
overview of the program stack and related mechanisms for stock OCaml.

In order to illustrate the layout of the stock OCaml stack, consider the
program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
main function |omain| installs two exception handlers |h1| and |h2| to handle
exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
which in turn calls back into OCaml function |c_to_ocaml|, which raises the
exception |E1|. OCaml supports raising exceptions in C as well as throwing
exceptions across external calls. Hence, the exception |E1| gets caught in the
handler |h1|, and |omain| returns |42|. The layout of the stack in native code
backend just before raising the exception in |c_to_ocaml| is illustrated in
Figure~\ref{fig:stock_stack}. The stack grows downwards.

OCaml uses the hardware stack as C, and hence the program stack has alternating
sequences of C and OCaml frames. Unlike C, OCaml does not create pointers into
OCaml frames. OCaml uses the hardware support for |call| and |return|
instructions for function calls and returns. OCaml does not perform explicit
stack overflow checks in code, and, just like C, relies on the guard page at
the end of the stack region to detect stack overflow. Stack overflow is
detected by a memory fault, and |Stack_overflow| exception is raised to unwind
the stack.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, a register each is
pinned for the \emph{allocation pointer} into the minor heap used for bump
pointer allocation, and the |Caml_state|, a table of global variables used by
the runtime. These registers are chosen to be C callee saved registers. This
makes external calls extremely fast in OCaml. If the external function does not
allocate in the OCaml heap, then the external functions can be called directly,
and no bookkeeping is necessary. For external functions which allocate in the
OCaml heap, the cached allocation pointer is saved to a field in the
|Caml_state| before the external call, and loaded back on return. Similarly,
callbacks into OCaml are also cheap. Callbacks into OCaml involves loading the
arguments in the right registers and calling the OCaml function. OCaml
callbacks are relatively common as the garbage collector (GC) finalisation
functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (|pc|)
of the handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|). The exception pointer is also updated to be the current stack
pointer (|sp|). This creates a linked-list of exception handler frames on the
stack as shown in Figure~\ref{fig:stock_stack}. Raising an exception simply
sets |sp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps to the |pc| of the
handler.

@@

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, which is the entry point for both the main OCaml code and
callbacks into OCaml, pushes an exception handler frame that forwards the
exception the innermost OCaml exception handler (|raised_exn_c| in
Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
very common to use them for \emph{local} control flow.

\subsection{Stack unwinding}
\ref{sec:unwind}

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting GC. For every call point in the program, OCaml compiler
emits the size of the frame and the set of all live registers in the frame
containing pointers into the heap. During a GC, the OCaml stack is walked and
the roots are marked.

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the following backtrace.
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\noindent which corresponds to the stack layout in
Figure~\ref{fig:stock_stack}.

@@

In this section, we refine the effect handler design for the OCaml use case.
The primary motivation for adding effect handlers to OCaml is to support
non-local control flow mechanism to implement asynchronous I/O in direct style,
generators and enable task parallelism. These use cases do not resume the
continuations more than once. Hence, our continuations are one-shot, and
resuming the continuation more than once raises |Invalid_argument _| exception.
It is well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96RepresentingControl}.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with _ -> (close_in ic; close_out oc)
\end{lstlisting}
|input_line| raises |End_of_file| exception on reaching the end of input.
Although not apparent, the code is written in a defensive style to handle other
exceptional cases such as the channels being closed externally. Both
|input_line| and |output_string| raise |Sys_error| exception if the channel is
closed. This function cleanly closes the channels on exceptional cases;
|close_*| functions do nothing if the channel is already closed.

@@

We can then use a top-level I/O handler similar to the introductory example
(\S\ref{sec:aio}) to discharge the I/O operations asynchronously. This handles
value return cases, but what about the exceptional cases |End_of_file| and
|Sys_error|? We introduce a |discontinue| primitive to resume a continuation by
raising an exception. In this example, on reaching the end of file, we would
discontinue the continuation of the |copy| function with |discontinue k End_of_file|,
which raises the exception at |input_line|, and the open channels
will be closed.

@@

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b

(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}

We introduce an extensible variant type~\cite{} |'a eff| of effect values,
which when performed using the |perform| primitive returns a |'a| value.
Constructors for the value of type |'a eff| are declared using the effect
declarations. For example, the declaration |effect E : string -> int| is
syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be |continue|d with a
suitably typed value or |discontinue|d with an exception.

>>>>

\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

@@

\newcommand{\mycaption}[1]{\vspace{-4mm}\caption{#1}}

@@

Effect Handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are defined by \emph{handlers} of the
effects. For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming.

@@

Common Lisp Conditions~\cite{Conditions} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision. If
the error is non-fatal, Conditions provide a \emph{restart} mechanism to resume
execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We can implement
this mechanism using effect handlers as shown below:

@@

On line 4, we declare an effect |DivBy0| which is parameterised by an integer,
which when performed returns result of division. If the denominator is 0, then
we perform |DivBy0| effect with the numerator value (line 12) so that the
context may decide the return value. We can get the |iOS| semantics by wrapping
the evaluator computation with the following handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}
If the evaluator returns a value, we simply return that. In case the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
other hand, we can get |gsearch| semantics by enclosing the evaluator in the
following handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
  if v = 0 then continue k Error
  else if v > 0 then continue k (Inf P)
  else continue k (Inf N)
\end{lstlisting}
The computation |gsearch(1/0)| returns positive infinity (|Inf P|).

@@

We can extend the calculator modularly with the ability to read user input
using the |Read| effect defined earlier as follows:
\begin{lstlisting}
type exp = ... | Inp
effect Read : in_channel -> string
let rec eval e = match e with
| ...
| Inp -> int_of_string(perform(Read stdin))
\end{lstlisting}
The |Read| effect can then be handled by another handler enclosing the
division-by-zero handler:
\begin{lstlisting}
let run e = match gsearch e with
| v -> v
| effect (Read chan) k ->
  register_async_input_line (chan, k);
  continue (get_ui_k ()) ()
\end{lstlisting}
\noindent which registers the read operation to be performed asynchronously
with the continuation as the callback. Subsequently, it resumes the
user-interface continuation so that the application remains responsive. Observe
that neither of the handlers needs to be aware of the other.

@@

OCaml, however, does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Koka}. If the compiler were to use CPS, then reifying
continuations in the source language is relatively straight-forward. But OCaml
does not use CPS IR, and changing the OCaml compiler to utilise CPS IR would be
an enormous undertaking. CPS IR would also affect the performance profile of
OCaml applications due to increased memory allocations as continuation closures
get allocated on the heap~\cite{Folklore}. Moreover, with CPS, stack unwinding
is meaningless and hence, we would lose compatibility with tools that inspect
the program stack. Hence, we choose not to use CPS translation and represent
continuations as call stacks.

The search for an expressive effect system that guarantees that all the effects
performed in the program are handled (\emph{effect safety}) in the presence of
advanced features such as polymorphism, modularity and generativity is an
active area of research~\cite{Leijen14, Biernacki19, Biernacki20,
Hillerstrom20}. We do not focus on this question in this paper, and our
implementation of effect handlers in OCaml does not guarantee effect safety.
For example, running |perform (Read stdin)| without an enclosing handler raises
|Unhandled| exception at the point of |perform|. We leave the question of
effect safety for future work.

@@

We motivate our effect handler design based on the following ideal
requirements:

\if{0}
\begin{enumerate}[label=R\arabic*]
  \item \textbf{Backwards compatibility} A well-behaved OCaml program does not
    break under OCaml extended with effect handlers. OCaml code that does not
    use effect handlers will pay negligible performance cost.
  \item \textbf{Tool compatibility} OCaml with effect handlers produce
    well-formed backtraces, and remains compatible with program analysis tools
    such as debuggers and profilers that inspect the stack using DWARF stack
    unwind tables.
  \item \textbf{Effect handler efficiency} In order to support
    highly-concurrent applications, the program must accommodate millions of
    continuations at the same time. Installing effect handlers, capturing and
    resuming continuations must be fast.
  \item \textbf{Forwards compatibility} While this work does not present an
    effect system, the design must be compatible with a future effect system
    for OCaml. As a cornerstone of modularity, we also want blocking I/O code
    to \emph{transparently} be made asynchronous with the help of effect
    handlers.
\end{enumerate}
\fi

\noindent \textbf{R1: Backwards compatibility.} A well-behaved OCaml program does
not break under OCaml extended with effect handlers. OCaml code that does not
use effect handlers will pay negligible performance cost.

\noindent \textbf{R2: Tool compatibility.} OCaml with effect handlers produce
well-formed backtraces, and remains compatible with program analysis tools such
as debuggers and profilers that inspect the stack using DWARF stack unwind
tables.

\noindent \textbf{R3: Effect handler efficiency.} In order to support highly-concurrent
applications, the program must accommodate millions of continuations at the
same time. Installing effect handlers, capturing and resuming continuations
must be fast.

\noindent \textbf{R4: Forwards compatibility.} While this work does not present an
effect system, the design must be compatible with a future effect system for
OCaml. As a cornerstone of modularity, we also want blocking I/O code to
\emph{transparently} be made asynchronous with the help of effect handlers.

The need to host millions of continuations at the same time rules out the use
of large contiguous stack space as in C for continuations. Instead, we resort
to using small initial stacks and growing the stacks on demand. As a result,
OCaml functions, irrespective of whether they use effect handlers, need to
perform stack overflow checks and external C functions, which do not have stack
overflow checks, must be performed on a separate system stack. Additionally, we
must generate DWARF stack unwind tables for stacks that may be non-contiguous.
In this work, we develop the compiler and runtime support for implementing
efficient effect handlers for OCaml that satisfy these requirements.

Our work is also timely. WebAssembly~\ref{} community group is actively
considering effect handlers as one of the mechanism for supporting concurrency,
asynchronous I/O and generators~\ref{WasmPropoal}. Project Loom~\cite{} is an
OpenJDK project which adds virtual threads and delimited continuations to Java.
Swift concurrency roadmap~\cite{} includes direct-style asynchronous
programming and structured concurrency as milestones. We believe that our
design choices will inform similar choices to be made in the other
industrial-strength languages.

@@

\begin{itemize}
  \item the design and implementation of one-shot effect handlers for OCaml.
    Our design retains OCaml's compatibility with program analysis tools that
    inspect the stack using DWARF unwind tables. We have validated our DWARF
    unwind tables using an automated validator tool~\cite{}.
  \item a formal operational semantics for the effect handler implementation in
    OCaml.
  \item extensive evaluation which shows that our implementation has negligible
    impact on code that does not use effect handlers, and serves as an
    efficient foundation for scalable concurrent and parallel programming.
\end{itemize}

We have implemented effect handlers in a multicore extension of the OCaml
programming language. We call our implementation ``Multicore OCaml'' to
distinguish it from stock OCaml.

The rest of the paper is organized as follows. We describe the details of the
OCaml program stack in the next section, and highlight the features that the
effect handler implementation must retain in order to provide backwards
compatibility. In \S\ref{sec:refine}, we refine the design further focussing on
the issues that come up when integrating effect handlers into a mainstream
systems language. \S\ref{sec:semantics} presents the formal operational
semantics for Multicore OCaml effect handlers. \S\ref{sec:impl} discusses the
compiler and the runtime system support for implementing effect handlers.
\S\ref{sec:eval} presents extensive performance evaluation of effect handlers
against the goals set in \S\ref{sec:req}. \S\ref{sec:related} and
\S\ref{sec:conc} present the related work and the conclusions.

@@

\begin{figure*}
\begin{minipage}{0.30\linewidth}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[numbers=left]
external ocaml_to_c
  : unit -> int = "ocaml_to_c"
exception E1
exception E2
let c_to_ocaml () = raise E1
let _ = Callback.register
    "c_to_ocaml" c_to_ocaml
let omain () =
  try (* h1 *)
    try (* h2 *) ocaml_to_c ()
    | with E2 -> 0
  | with E1 -> 42;;
let _ = assert (omain () = 42)
    \end{lstlisting}
    \vspace{-2mm}
    \subcaption{\texttt{meander.ml}}
    \label{code:meander_ml}
  \end{minipage}
  \begin{minipage}{\linewidth}
    \begin{lstlisting}[language=c,numbers=left]
#include <caml/mlvalues.h>
#include <caml/callback.h>

value ocaml_to_c (value unit) {
  caml_callback(*caml_named_value
    ("c_to_ocaml"), Val_unit);
  return Val_int(0);
}
    \end{lstlisting}
    \vspace{-4mm}
    \subcaption{\texttt{meander.c}}
    \label{code:meander_c}
  \end{minipage}
\end{minipage}
%
\begin{minipage}{0.30\linewidth}
  \centering
  \includegraphics[scale=0.46]{figures/stock_stack}
  \subcaption{Stack layout before raise \texttt{E1}.}
  \label{fig:stock_stack}
\end{minipage}
%
\begin{minipage}{0.39\linewidth}
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:13
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\vspace{-2mm}
\subcaption{\texttt{gdb} backtrace before raise \texttt{E1}.}
\label{code:gdb_backtrace}
\end{minipage}
  \mycaption{Program stack on stock OCaml.}
\end{figure*}

The main challenge in implementing effect handlers in Multicore OCaml is
managing the program stack. To this end, in this section, we provide an
overview of the program stack and related mechanisms in stock OCaml.

In order to illustrate the layout of the stock OCaml stack, consider the
program in Figures~\ref{code:meander_ml} and~\ref{code:meander_c}. The OCaml
main function |omain| installs two exception handlers |h1| and |h2| to handle
exceptions |E1| and |E2|. |omain| calls the external C function |ocaml_to_c|,
which in turn calls back into OCaml function |c_to_ocaml|, which raises the
exception |E1|. OCaml supports raising exceptions in C as well as throwing
exceptions across external calls. Hence, the exception |E1| gets caught in the
handler |h1|, and |omain| returns |42|. The layout of the stack in native code
backend just before raising the exception in |c_to_ocaml| is illustrated in
Figure~\ref{fig:stock_stack}. Note that the stack grows downwards.

OCaml uses the same hardware stack as C, and hence the program stack has
alternating sequences of C and OCaml frames. However, unlike C, OCaml does not
create pointers into OCaml frames. OCaml uses the hardware support for |call|
and |return| instructions for function calls and returns. OCaml does not
perform explicit stack overflow checks in code, and, just like C, relies on the
guard page at the end of the stack region to detect stack overflow. Stack
overflow is detected by a memory fault, and |Stack_overflow| exception is
raised to unwind the stack.

@@

OCaml does not use the C calling convention. In particular, there are no
callee-saved registers in OCaml. On the |x86-64| backend, the C callee-saved
registers, |r15| holds the \emph{allocation pointer} into the minor heap used
for bump pointer allocation, and |r14| holds the |Caml_state|, a table of
global variables used by the runtime. This makes external calls extremely fast
in OCaml. If the external function does not allocate in the OCaml heap, then
the external functions can be called directly, and no bookkeeping is necessary.
For external functions which allocate in the OCaml heap, the cached allocation
pointer is saved to a field in the |Caml_state| before the external call, and
loaded back on return. Similarly, callbacks into OCaml are also cheap.
Callbacks into OCaml involves loading the arguments in the right registers and
calling the OCaml function. OCaml callbacks are relatively common as the
garbage collector (GC) finalisation functions are executed as OCaml callbacks.

@@

The lack of callee-saved registers also makes exception handling fast as there
are no callee-saved registers to be restored when the exception handler runs.
Hence, installing an exception handler simply pushes the program counter (|pc|)
of the handler and the current exception pointer (|exn_ptr| -- a field in
|Caml_state|). At this point, the exception pointer is updated to be the
current stack pointer (|rsp|). This creates a linked-list of exception handler
frames on the stack as shown in Figure~\ref{fig:stock_stack}. Raising an
exception simply sets |rsp| to |exn_ptr|, loads the saved |exn_ptr|, and jumps
to the |pc| of the handler.

@@

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, which is the entry point for both the main OCaml code and
callbacks into OCaml, pushes an exception handler frame that forwards the
exception to the innermost OCaml exception handler (|raise_exn_c| in
Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
very common to use them for \emph{local} control flow.

\subsection{Stack unwinding}
\label{sec:unwind}

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting GC. For every call point in the program, OCaml compiler
emits the size of the frame and the set of all live registers in the frame
containing pointers into the heap. During a GC, the OCaml stack is walked and
the roots are marked skipping over the C frames.

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the backtrace in Figuere~\ref{code:gdb_backtrace} which
corresponds to the stack layout in Figure~\ref{fig:stock_stack}.

@@

In this section, we refine the effect handler design for the OCaml use case.
The primary motivation for adding effect handlers to OCaml is to support
non-local control flow mechanism to implement asynchronous I/O in direct style,
generators and enable task parallelism. These use cases do not resume the
continuations more than once. Hence, our continuations are one-shot, and
resuming the continuation more than once raises |Invalid_argument| exception.
It is well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96RepresentingControl}.

@@

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17}. Consider the following function that uses blocking I/O
functions from the OCaml standard library to copy data from the input channel
|ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with _ -> (close_in ic; close_out oc)
\end{lstlisting}

The function |input_line| raises |End_of_file| exception on reaching the end of input.
Although not apparent, the code is written in a defensive style to handle other
exceptional cases such as the channels being closed externally. Both
|input_line| and |output_string| raise |Sys_error| exception if the channel is
closed. This function cleanly closes the channels on exceptional cases;
|close_*| functions do nothing if the channel is already closed.

@@

We can then use a top-level I/O handler similar to the introductory example
(\S\ref{sec:aio}) to discharge the I/O operations asynchronously and resumes
with the result. This handles value return cases, but what about the
exceptional cases |End_of_file| and |Sys_error|? We introduce a |discontinue|
primitive to resume a continuation by raising an exception. In this example, on
reaching the end of file, we would discontinue the continuation of the |copy|
function with |discontinue k End_of_file|, which raises the exception at
|input_line|, and the open channels will be closed.

@@

The static semantics of effect handlers in OCaml is captured succinctly by its
API:
\begin{lstlisting}
type 'a eff = ..
type ('a,'b) continuation
val perform: 'a eff -> 'a
val continue: ('a,'b) continuation -> 'a -> 'b
val discontinue: ('a,'b) continuation -> exn -> 'b
(* Internal API *)
type 'a comp = unit -> 'a
type ('a,'b) handler =
{retc: 'a -> 'b;
 effc: 'c.'c eff -> ('c,'b)continuation -> 'b; }
val match_with: 'a comp -> ('a,'b) handler -> 'b
\end{lstlisting}
We introduce an extensible variant type~\cite{} |'a eff| of effect values,
which when performed using the |perform| primitive returns a |'a| value.
Constructors for the value of type |'a eff| are declared using the effect
declarations. For example, the declaration |effect E : string -> int| is
syntactic sugar for adding a new constructor to the variant type
|type _ eff += E : string -> int eff|. We introduce the type
|('a,'b) continuation| of delimited continuations which expect a |'a| value for
resumption and return |'b| value. The continuations may be |continue|d with a
suitably typed value or |discontinue|d with an exception.
